---
title: "HW2 - P8105"
output: github_document
date: "2025-09-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# HW 2
## R Markdown
### Subtitle 
#### Tester
##### Tester 3
###### Tester 4
Random Text

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

## Problem 1
Data to be used: `pols-month.csv`, `unemployment.csv`, and `snp.csv`. 
Goal: Merge these into a single data frame using year and month as keys across datasets.

### Step 0: Import and View Data:
Import each of the datasets and view what we're working with. 
```{r}
pols_month_df =
  read_csv("fivethirtyeight_datasets/pols-month.csv") |> 
  janitor::clean_names() 

#When imported everything is already dbl meaning it is a double numeric variable (aka we don't need to include a line like --> na = c("NA", ".", "") that converts missing values)

snp_df =
  read_csv("fivethirtyeight_datasets/snp.csv") |> 
  janitor::clean_names() 


unemployment_df =
  read_csv("fivethirtyeight_datasets/unemployment.csv") |> 
  janitor::clean_names() 
```


### Step 1: Clean `pols-month.csv`
Use separate() to break up the variable mon into integer variables year, month, and day; 
Replace month number with month name; 
Create a president variable taking values gop and dem,
Remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols_month_df =  
 separate(pols_month_df, mon, c("year","month","day"),"-", convert = TRUE) |> 
  mutate(
    month = month.name[month],
    president = case_when(
      prez_dem == 1 ~ "dem",
      prez_gop == 1 ~ "gop"
    )) |> 
  select(-prez_dem, -prez_gop, -day)

 #Found this shortcut online. Instead of mutate(month = month.name[month]), you could also use mutate(month = case_when(month == 1 ~ "January",....
```

### Step 2: Clean `snp.csv`
Use a similar process to the above. 
For consistency across datasets, arrange according to year and month, 
Organize so that year and month are the leading columns.

```{r}
snp_df =  
 separate(snp_df, date, c("month","day","year"),"/", convert = TRUE) |> 
  mutate(
    month = month.name[month],
    year = if_else(year <= 15, year + 2000, year + 1900)) |> 
  select(-day) |> 
  relocate(year)

```


### Step 3: Tidy `unemployment.csv`
Tidy the unemployment data so that it can be merged with the previous datasets. 
Switch it from “wide” to “long” format --> ensure key variables have the same name and that key variables take the same values.


```{r}
unemployment_df <-
unemployment_df |>   
  pivot_longer( 
    cols = jan:dec,
    names_to = "month",
    values_to = "unemployment_rate") |> 
  mutate(month = tolower(month),
        month = case_when( 
                            month == "jan" ~ "January",
                            month == "feb" ~ "February",
                            month == "mar" ~ "March",
                            month == "apr" ~ "April",
                            month == "may" ~ "May",
                            month == "jun" ~ "June",
                            month == "jul" ~ "July",
                            month == "aug" ~ "August",
                            month == "sep" ~ "September",
                            month == "oct" ~ "October",
                            month == "nov" ~ "November",
                            month == "dec" ~ "December"))

 

```



### Step 4: Join Data Sets
Merge `snp.csv` INTO `pols-month.csv`,
Merge `unemployment.csv` into above result.

```{r}

snps_pols_df = 
  left_join(pols_month_df, snp_df, by = c("year","month"))


unemployment_snps_pols_df = 
  left_join(snps_pols_df, unemployment_df, by = c("year","month"))

```



### Step 5: Analyse Results
Write a short paragraph about these datasets. 
Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).


I DONT WANT TO LOL



#### Additional Information & Side Notes:
Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.

-------------------------------------------------------------------------

## Problem 2
This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website.

### Step 1: Read and clean the `Mr. Trash Wheel sheet`:
specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
use reasonable variable names
omit rows that do not include dumpster-specific data
round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
#omit first row
#second row is header
#row 654 is blank and 655 is total of column values
#Homes column includes 0s and blanks


mr_trash_df <-
  read_excel("Trashwheel_Collection_Data.xlsx",  sheet = "Mr. Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  select(-starts_with("x"), -homes_powered) |> 
  filter(!is.na(dumpster)) |> 
  mutate(sports_balls = as.integer(round(sports_balls)),
         sheet = "mr_trash",
         year = as.double(year))




#filter(!is.na(dumpster)) removes the last two rows which includes a blank row and the summary total row
```




### Step 2: Import, Clean, and Merge 
Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r}

prof_trash_df <-
  read_excel("Trashwheel_Collection_Data.xlsx",  sheet = "Professor Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  select(-homes_powered) |> 
  filter(!is.na(dumpster)) |> 
  mutate(sheet = "prof_trash")


gwyn_trash_df <-
  read_excel("Trashwheel_Collection_Data.xlsx",  sheet = "Gwynnda Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  select(-homes_powered) |> 
  filter(!is.na(dumpster)) |> 
  mutate(sheet = "gwyn_trash")


#Combine all three datasets 

all_trash_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwyn_trash_df)

```


### Step 3: Discuss Findings
Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?



-----------------------------------------------------------------------------

## Problem 3

Home and rental prices have generally increased over the last decade. Zillow, a popular website used to search for homes for sale or rent, is uniquely positioned to provide insights into trends in the real estate market. In response to broad interest, the company releases data for research. In this project, we’ll look at the Zillow Observed Rent Index (ZORI) in New York City between January 2015 and August 2024.

NYC is divided into five boroughs. Each of these boroughs is it’s own county, and in some cases the borough name and county name differ; for example, Manhattan is New York County. Moreover, boroughs are divided into neighborhoods. Rental price data provided by Zillow does not include information neighborhoods within boroughs, but can be accessed separately.

### Step 1: Create a single dataset
Create a single, well-organized dataset with all the information contained in these data files. 
To that end: import, clean, tidy, and otherwise wrangle each of these datasets; 
check for completeness and correctness across datasets (e.g. by viewing individual datasets and monitoring warning messages); merge to create a single, final dataset; 
and organize this so that variables and observations are in meaningful orders.






### Step 2: Describe Results

Briefly describe the resulting tidy dataset. How many total observations exist? How many unique ZIP codes are included, and how many unique neighborhoods?






### Step 3: Zipcodes and Graphics
Which ZIP codes appear in the ZIP code dataset but not in the Zillow Rental Price dataset? Using a few illustrative examples discuss why these ZIP codes might be excluded from the Zillow dataset.





### Covid Fluctuations

Rental prices fluctuated dramatically during the COVID-19 pandemic. For all available ZIP codes, compare rental prices in January 2021 to prices in January 2020. Make a table that shows the 10 ZIP codes (along with the borough and neighborhood) with largest drop in price from January 2020 to 2021. Comment.








